---
title: "Chicago Bulls Team Analysis"
author: "Ryan Ahchee"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

### Description
The Chicago Bulls are a team that competes in the NBA (National Basketball Association). In the most recent NBA season they placed 27th out of 30 (based on win-loss record). 
The task by the general manager of Chicago Bulls is to find the best five starting players (one from each position) your team can afford but still having the ability to fill a full team roster. For the purpose of this task, we are focused on finding five starting players and can include players already on the Chicago Bulls roster. 
The restriction for this task are that the team's budget for player contracts next season is 118 million, ranked 26th out of 30. For context, the team with the highest payroll budget is Portland with 148 million, while the best performing team was Milwaukee Bucks (who clinched the best league record in 2018-19) with $131 million. 

2. To demonstrate the magic of Rmarkdown and convince you to always use Rmarkdown to document your analysis

### Background on Key Basketball Metrics

### Aim of Project
1. Explore basic data analysis techniques for exploring, describing, and summarising data
2. Provide recommendations to to General Manager on best recruitment players for Chicago Bulls in order to remove  

## Load required packages

```{r packages, message=FALSE}
library(tidyverse) # include all required packages at the start
```

## Data source

The data used in this document contains match data form the National Basketball League for season 2018-19. 

It has been sourced from the below sources.
1. 2018-19_nba_player-statistics.csv :  sourced from [www.basketball-reference.com] basketball-reference.com  
2. 2018-19_nba_player-salaries.csv :  sourced from [www.hoopshype.com/salaries]hoopshype.com/salaries 

3. 2019-20_nba_team-payroll.csv : sourced from [www.hoopshype.com/salaries]hoopshype.com/salaries

4. 2018-19_nba_team-statistics_1.csv : sourced from [www.basketball-reference.com]basketball-reference.com

5. 2018-19_nba_team-statistics_2.csv : sourced from [www.basketball-reference.com]basketball-reference.com

A description of all data variable descriptions can be found [https://uclearn.canberra.edu.au/courses/9531/pages/data-description-reproducible-data-analysis-project]here
## Read data

Read in the `2018-19_nhl_match-data.csv` file using the `read_csv()` function from the `readr` package. 

```{r read_data, message=FALSE} 
# read in the data
# data dir should be in the working dir
df <- read_csv("data/raw/2018-19_nba_player-statistics.csv")
```

## Checking the data

Check the structure of the data file using `str()`:

```{r structure}
str(df)
```

Check the first 6 rows of the data file using `head()`

```{r head}
knitr::kable(head(df))
```

Check the last 6 rows of the data file using `tail()`

```{r tail}
tail(df)
```

Check for missing values using:

```{r na}
sum(is.na(df)) # count the missing values
head(which(is.na(df), arr.ind = TRUE)) # find where the missing values are (showing only first 6 rows)
```

The missing values all appear in the `Time` variable when the game finished in regular time (i.e. it did not go to overtime or a shoot-out). 

We can impute the missing values with the string "REG" for regular time using the`replace_na()` function from the dplyr package:

```{r replace_na}
df <- replace_na(df, list(Time = "REG"))
sum(is.na(df)) # check for missing values again
```

## Exploratory analysis

Let's create some summary statistics per team (e.g. `Tm` variable)

```{r summarising}
team_outcomes <- 
  df %>%
    count(Tm, Outcome) %>%
    group_by(Tm) %>%
    mutate(prob = round(prop.table(n), 2))

team_outcomes
```

We can create a barplot from the summarised data above using `ggplot2` to visualise frequencies and proportions

```{r barplot}
team_outcomes %>%
  ggplot(aes(x = Tm, y = prob, fill = Outcome)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90))
```
